---
title: "Enrichment analysis"
author: "Gema Castillo García"
date: "20/05/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = 'C:/Users/Acerolo/Desktop/GitHub/TFM')
knitr::opts_knit$set(root.dir = 'C:/Users/gemac/OneDrive - Fundación Universitaria San Pablo CEU/UPM/Máster/Semestre II/TFM/scripts GitHub')
```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(clusterProfiler)
library(org.At.tair.db)
library(biomartr) #
library(enrichplot)
library(ggplot2)
library(pathview)

keytypes(org.At.tair.db) #checking ID types supported by org.At.tair.db (TAIR is accepted)
```

A gene set is an unordered collection of genes that are functionally related.


# Gene Ontology (GO)

GO comprises three ontologies: molecular function (MF), biological process (BP), and cellular component (CC). In this case, I am interested on the second one (pathways and larger processes made up of the activities of multiple gene products).



# HB21ind DE genes #


### GO over-representation analysis of common DE genes in HBs
```{r}
# Test which biological processes (BP) are over-represented in common down DE genes in HBs

DE_DN_HBs_enrichedGO_Infmax = enrichGO(gene         = common_DOWN_genes,
                                      OrgDb         = org.At.tair.db,
                                      keyType       = "TAIR",
                                      ont           = "BP",
                                      minGSSize     = 10,
                                      maxGSSize     = Inf,
                                      pAdjustMethod = "BH",
                                      pvalueCutoff  = 0.05,
                                      qvalueCutoff  = 1,
                                      readable      = TRUE) #%>% simplify()

# Save filtered GO terms in a table
DE_DN_HBs_enrichedGO_Infmax@result %>% filter(p.adjust<0.05) %>% write.table(.,
            file = "./R_output/GO_terms_enrichment_analysis/common_HBs_DOWN_DE_list_GO_terms.txt",
            row.names = F, sep = "\t", quote = F, na = "")




# Test which biological processes (BP) are over-represented in common up DE genes in HBs
DE_UP_HBs_enrichedGO_Infmax = enrichGO(gene         = common_UP_genes,
                                      OrgDb         = org.At.tair.db,
                                      keyType       = "TAIR",
                                      ont           = "BP",
                                      minGSSize     = 10,
                                      maxGSSize     = Inf,
                                      pAdjustMethod = "BH",
                                      pvalueCutoff  = 0.05,
                                      qvalueCutoff  = 1,
                                      readable      = TRUE) #%>% simplify()

# Save filtered GO terms in a table
DE_UP_HBs_enrichedGO_Infmax@result %>% filter(p.adjust<0.05) %>% write.table(.,
            file = "./R_output/GO_terms_enrichment_analysis/common_HBs_UP_DE_list_GO_terms.txt",
            row.names = F, sep = "\t", quote = F, na = "")
```



### GO over-representation analysis of common UP DTs of HBs
```{r}
# Test which biological processes (BP) are over-represented in common down DE genes in HBs

UP_DT_HBs_enrichedGO_Infmax = enrichGO(gene         = common_UP_DT_HBs,
                                      OrgDb         = org.At.tair.db,
                                      keyType       = "TAIR",
                                      ont           = "BP",
                                      minGSSize     = 10,
                                      maxGSSize     = Inf,
                                      pAdjustMethod = "BH",
                                      pvalueCutoff  = 0.05,
                                      qvalueCutoff  = 1,
                                      readable      = TRUE) #%>% simplify()

# Save filtered GO terms in a table
UP_DT_HBs_enrichedGO_Infmax@result %>% filter(p.adjust<0.05) %>% write.table(.,
            file = "./R_output/GO_terms_enrichment_analysis/HBs_common_UP_DT_list_GO_terms.txt",
            row.names = F, sep = "\t", quote = F, na = "")

```



### GO over-representation analysis of common BF DT of HBs
```{r}
# Test which biological processes (BP) are over-represented in common down DE genes in HBs

BF_DT_HBs_enrichedGO_Infmax = enrichGO(gene         = HBs_BF_DT$Geneid,
                                      OrgDb         = org.At.tair.db,
                                      keyType       = "TAIR",
                                      ont           = "BP",
                                      minGSSize     = 10,
                                      maxGSSize     = Inf,
                                      pAdjustMethod = "BH",
                                      pvalueCutoff  = 0.05,
                                      qvalueCutoff  = 1,
                                      readable      = TRUE) #%>% simplify()

# Save filtered GO terms in a table
BF_DT_HBs_enrichedGO_Infmax@result %>% filter(p.adjust<0.05) %>% write.table(.,
            file = "./R_output/GO_terms_enrichment_analysis/HBs_BF_DT_list_GO_terms.txt",
            row.names = F, sep = "\t", quote = F, na = "")

```