---
title: "Enrichment analysis"
author: "Gema Castillo García"
date: "20/05/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = 'C:/Users/Acerolo/Desktop/GitHub/TFM')
knitr::opts_knit$set(root.dir = 'C:/Users/gemac/OneDrive - Fundación Universitaria San Pablo CEU/UPM/Máster/Semestre II/TFM/scripts GitHub')
```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(clusterProfiler)
library(org.At.tair.db)
library(biomartr) #
library(enrichplot)
library(ggplot2)
library(pathview)

keytypes(org.At.tair.db) #checking ID types supported by org.At.tair.db (TAIR is accepted)
```

A gene set is an unordered collection of genes that are functionally related.


# Gene Ontology (GO)

GO comprises three ontologies: molecular function (MF), biological process (BP), and cellular component (CC). In this case, I am interested on the second one (pathways and larger processes made up of the activities of multiple gene products).



# wt_ABA vs wt_ctr #

Some of the following functions require a ranked gene list, which contains three features:

    numeric vector: fold change (FC) or other type of numerical variable
    named vector: every FC has the corresponding gene ID
    sorted vector: FCs should be sorted in decreasing order
    
```{r}
# Create a df with UP-regulated genes in wt_aba_vs_wt_ctr and their FC
aba_wt_UP_DE_list = edgeR_results_ABA$list_by_contrast$wt_aba_vs_wt_ctr %>% filter(logFC_wt_aba_vs_wt_ctr > 1 & FDR_wt_aba_vs_wt_ctr < 0.05)
row.names(aba_wt_UP_DE_list) = aba_wt_UP_DE_list$Geneid

#Creating a geneList with -log10(FC) values of genes with FDR < 0.05
  ## feature 1: numeric vector
  aba_wt_UP_geneList_FC = aba_wt_UP_DE_list$logFC_wt_aba_vs_wt_ctr
  
  ## feature 2: named vector
  names(aba_wt_UP_geneList_FC) = as.character(row.names(aba_wt_UP_DE_list))
  
  ## feature 3: sorted vector
  aba_wt_UP_geneList_FC = sort(aba_wt_UP_geneList_FC, decreasing = TRUE)
  
  
  
  
# Create a df with DOWN-regulated genes in wt_aba_vs_wt_ctr
aba_wt_DN_DE_list = edgeR_results_ABA$list_by_contrast$wt_aba_vs_wt_ctr %>% filter(logFC_wt_aba_vs_wt_ctr < 1 & FDR_wt_aba_vs_wt_ctr < 0.05)
row.names(aba_wt_DN_DE_list) = aba_wt_DN_DE_list$Geneid

#Creating a geneList with -log10(FC) values of genes with FDR < 0.05
  ## feature 1: numeric vector
  aba_wt_DN_geneList_FC = aba_wt_DN_DE_list$logFC_wt_aba_vs_wt_ctr
  
  ## feature 2: named vector
  names(aba_wt_DN_geneList_FC) = as.character(row.names(aba_wt_DN_DE_list))
  
  ## feature 3: sorted vector
  aba_wt_DN_geneList_FC = sort(aba_wt_DN_geneList_FC, decreasing = TRUE)
```



### GO over-representation analysis of UP-rgeulated ABA-responsive genes in wt
```{r}
# Test which biological processes (BP) are over-represented in UP DE wt_aba_vs_wt_ctr

aba_wt_UP_DE_enrichedGO_250max = enrichGO(gene          = names(aba_wt_UP_geneList_FC),
                                          OrgDb         = org.At.tair.db,
                                          keyType       = "TAIR",
                                          ont           = "BP",
                                          minGSSize     = 10,
                                          maxGSSize     = 250,
                                          pAdjustMethod = "BH",
                                          pvalueCutoff  = 0.05,
                                          qvalueCutoff  = 1,
                                          readable      = TRUE) #%>% simplify()

aba_wt_UP_DE_enrichedGO_500max = enrichGO(gene          = names(aba_wt_UP_geneList_FC),
                                          OrgDb         = org.At.tair.db,
                                          keyType       = "TAIR",
                                          ont           = "BP",
                                          minGSSize     = 10,
                                          maxGSSize     = 500,
                                          pAdjustMethod = "BH",
                                          pvalueCutoff  = 0.05,
                                          qvalueCutoff  = 1,
                                          readable      = TRUE) #%>% simplify()

aba_wt_UP_DE_enrichedGO_1000max = enrichGO(gene          = names(aba_wt_UP_geneList_FC),
                                           OrgDb         = org.At.tair.db,
                                           keyType       = "TAIR",
                                           ont           = "BP",
                                           minGSSize     = 10,
                                           maxGSSize     = 1000,
                                           pAdjustMethod = "BH",
                                           pvalueCutoff  = 0.05,
                                           qvalueCutoff  = 1,
                                           readable      = TRUE) #%>% simplify()

aba_wt_UP_DE_enrichedGO_Infmax = enrichGO(gene          = names(aba_wt_UP_geneList_FC),
                                          OrgDb         = org.At.tair.db,
                                          keyType       = "TAIR",
                                          ont           = "BP",
                                          minGSSize     = 10,
                                          maxGSSize     = Inf,
                                          pAdjustMethod = "BH",
                                          pvalueCutoff  = 0.05,
                                          qvalueCutoff  = 1,
                                          readable      = TRUE) #%>% simplify()

# Save filtered GO terms in a table
aba_wt_UP_DE_enrichedGO_Infmax@result %>% filter(p.adjust<0.05) %>% write.table(.,
            file = "./R_output/GO_terms_enrichment_analysis/aba_wt_UP_DE_list_GO_terms.txt",
            row.names = F, sep = "\t", quote = F, na = "")

```

### Visualization of enriched GO terms
```{r}
aba_wt_UP_cnetplot_250maxGSS = cnetplot(aba_wt_UP_DE_enrichedGO_250max, categorySize="p.adjust",
                                   foldChange=aba_wt_UP_geneList_FC) # category netplot with FC values
aba_wt_UP_cnetplot_500maxGSS = cnetplot(aba_wt_UP_DE_enrichedGO_500max, categorySize="p.adjust",
                                   foldChange=aba_wt_UP_geneList_FC) # category netplot with FC values
aba_wt_UP_cnetplot_1000maxGSS = cnetplot(aba_wt_UP_DE_enrichedGO_1000max, categorySize="p.adjust", 
                                    foldChange=aba_wt_UP_geneList_FC) # category netplot with FC values

# Saving Gene-Concept Networks
png("./R_output/GO_terms_enrichment_analysis/aba_wt_UP_DE_Gene-Concept_Net_250maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
aba_wt_UP_cnetplot_250maxGSS

png("./R_output/GO_terms_enrichment_analysis/aba_wt_UP_DE_Gene-Concept_Net_500maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
aba_wt_UP_cnetplot_500maxGSS
dev.off()

png("./R_output/GO_terms_enrichment_analysis/aba_wt_UP_DE_Gene-Concept_Net_1000maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
aba_wt_UP_cnetplot_1000maxGSS
dev.off()
```


### GO over-representation analysis of DOWN-rgeulated ABA-responsive genes in wt
```{r}
# Test which biological processes (BP) are over-represented in DN DE wt_aba_vs_wt_ctr

aba_wt_DN_DE_enrichedGO_500max = enrichGO(gene          = names(aba_wt_DN_geneList_FC),
                                          OrgDb         = org.At.tair.db,
                                          keyType       = "TAIR",
                                          ont           = "BP",
                                          minGSSize     = 10,
                                          maxGSSize     = 500,
                                          pAdjustMethod = "BH",
                                          pvalueCutoff  = 0.05,
                                          qvalueCutoff  = 1,
                                          readable      = TRUE) #%>% simplify()

aba_wt_DN_DE_enrichedGO_1000max = enrichGO(gene          = names(aba_wt_DN_geneList_FC),
                                          OrgDb         = org.At.tair.db,
                                          keyType       = "TAIR",
                                          ont           = "BP",
                                          minGSSize     = 10,
                                          maxGSSize     = 1000,
                                          pAdjustMethod = "BH",
                                          pvalueCutoff  = 0.05,
                                          qvalueCutoff  = 1,
                                          readable      = TRUE) #%>% simplify()



aba_wt_DN_DE_enrichedGO_Infmax = enrichGO(gene          = names(aba_wt_DN_geneList_FC),
                                          OrgDb         = org.At.tair.db,
                                          keyType       = "TAIR",
                                          ont           = "BP",
                                          minGSSize     = 10,
                                          maxGSSize     = Inf,
                                          pAdjustMethod = "BH",
                                          pvalueCutoff  = 0.05,
                                          qvalueCutoff  = 1,
                                          readable      = TRUE) #%>% simplify()

# Save filtered GO terms in a table
aba_wt_DN_DE_enrichedGO_Infmax@result %>% filter(p.adjust<0.05) %>% write.table(.,
            file = "./R_output/GO_terms_enrichment_analysis/aba_wt_DN_DE_list_GO_terms.txt",
            row.names = F, sep = "\t", quote = F, na = "")

```

### Visualization of enriched GO terms
```{r}
aba_wt_DN_cnetplot_500maxGSS = cnetplot(aba_wt_DN_DE_enrichedGO_500max, categorySize="p.adjust",
                                   foldChange=aba_wt_DN_geneList_FC) # category netplot with FC values
aba_wt_DN_cnetplot_1000maxGSS = cnetplot(aba_wt_DN_DE_enrichedGO_1000max, categorySize="p.adjust", 
                                    foldChange=aba_wt_DN_geneList_FC) # category netplot with FC values

# Saving Gene-Concept Networks
png("./R_output/GO_terms_enrichment_analysis/aba_wt_DN_DE_Gene-Concept_Net_500maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
aba_wt_DN_cnetplot_500maxGSS
dev.off()

png("./R_output/GO_terms_enrichment_analysis/aba_wt_DN_DE_Gene-Concept_Net_1000maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
aba_wt_DN_cnetplot_1000maxGSS
dev.off()
```