---
title: "Enrichment analysis"
author: "Gema Castillo García"
date: "20/05/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = 'C:/Users/Acerolo/Desktop/GitHub/TFM')
knitr::opts_knit$set(root.dir = 'C:/Users/gemac/OneDrive - Fundación Universitaria San Pablo CEU/UPM/Máster/Semestre II/TFM/scripts GitHub')
```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(clusterProfiler)
library(org.At.tair.db)
library(biomartr) #
library(enrichplot)
library(ggplot2)
library(pathview)

keytypes(org.At.tair.db) #checking ID types supported by org.At.tair.db (TAIR is accepted)
```

A gene set is an unordered collection of genes that are functionally related.


# Gene Ontology (GO)

GO comprises three ontologies: molecular function (MF), biological process (BP), and cellular component (CC). In this case, I am interested on the second one (pathways and larger processes made up of the activities of multiple gene products).



# GFP:BRC1ind DE genes #

Some of the following functions require a ranked gene list, which contains three features:

    numeric vector: fold change (FC) or other type of numerical variable
    named vector: every FC has the corresponding gene ID
    sorted vector: FCs should be sorted in decreasing order
```{r}
row.names(BRC1_DE_list) = BRC1_DE_list$Geneid

#Creating a geneList with -log10(FC) values of genes with FDR < 0.05
  ## feature 1: numeric vector
  BRC1_DE_geneList_FC = BRC1_DE_list$logFC_GFP_BRC1ind_vs_brc1
  
  ## feature 2: named vector
  names(BRC1_DE_geneList_FC) = as.character(row.names(BRC1_DE_list))
  
  ## feature 3: sorted vector
  BRC1_DE_geneList_FC = sort(BRC1_DE_geneList_FC, decreasing = TRUE)
```

### GO over-representation analysis of DE genes in BRC1ind
```{r}
# Test which biological processes (BP) are over-represented in BF DT of BRC1

DE_BRC1_enrichedGO_250max = enrichGO(gene          = row.names(BRC1_DE_list),
                                     OrgDb         = org.At.tair.db,
                                     keyType       = "TAIR",
                                     ont           = "BP",
                                     minGSSize     = 10,
                                     maxGSSize     = 250,
                                     pAdjustMethod = "BH",
                                     pvalueCutoff  = 0.05,
                                     qvalueCutoff  = 1,
                                     readable      = TRUE) #%>% simplify()

DE_BRC1_enrichedGO_500max = enrichGO(gene          = row.names(BRC1_DE_list),
                                     OrgDb         = org.At.tair.db,
                                     keyType       = "TAIR",
                                     ont           = "BP",
                                     minGSSize     = 10,
                                     maxGSSize     = 500,
                                     pAdjustMethod = "BH",
                                     pvalueCutoff  = 0.05,
                                     qvalueCutoff  = 1,
                                     readable      = TRUE) #%>% simplify()

DE_BRC1_enrichedGO_1000max = enrichGO(gene          = row.names(BRC1_DE_list),
                                      OrgDb         = org.At.tair.db,
                                      keyType       = "TAIR",
                                      ont           = "BP",
                                      minGSSize     = 10,
                                      maxGSSize     = 1000,
                                      pAdjustMethod = "BH",
                                      pvalueCutoff  = 0.05,
                                      qvalueCutoff  = 1,
                                      readable      = TRUE) #%>% simplify()

# # Save filtered GO terms in a table
# DE_BRC1_enrichedGO_1000max@result %>% filter(p.adjust<0.05) %>% write.table(.,
#             file = "./R_output/GO_terms_enrichment_analysis/BRC1_DE_list_GO_terms.txt",
#             row.names = F, sep = "\t", quote = F, na = "")
```

### Visualization of enriched GO terms
```{r}
#goplot(BF_DT_BRC1_enrichedGO) # directed acyclic graph
DE_BRC1_cnetplot_250maxGSS = cnetplot(DE_BRC1_enrichedGO_250max, categorySize="p.adjust",
                                   foldChange=BRC1_DE_geneList_FC) # category netplot with FC values
DE_BRC1_cnetplot_500maxGSS = cnetplot(DE_BRC1_enrichedGO_500max, categorySize="p.adjust",
                                   foldChange=BRC1_DE_geneList_FC) # category netplot with FC values
DE_BRC1_cnetplot_1000maxGSS = cnetplot(DE_BRC1_enrichedGO_1000max, categorySize="p.adjust", 
                                    foldChange=BRC1_DE_geneList_FC) # category netplot with FC values

# Saving Gene-Concept Networks
png("./R_output/GO_terms_enrichment_analysis/BRC1ind_DE_Gene-Concept_Net_250maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
DE_BRC1_cnetplot_250maxGSS
dev.off()

png("./R_output/GO_terms_enrichment_analysis/BRC1ind_DE_Gene-Concept_Net_500maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
DE_BRC1_cnetplot_500maxGSS
dev.off()

png("./R_output/GO_terms_enrichment_analysis/BRC1ind_DE_Gene-Concept_Net_1000maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
DE_BRC1_cnetplot_1000maxGSS
dev.off()
```



# GFP:BRC1ind UP DE genes #
```{r}
BRC1_UP_DE_list = BRC1_DE_list %>% filter(logFC_GFP_BRC1ind_vs_brc1>1)

#Creating a geneList with -log10(FC) values of genes with FDR < 0.05
  ## feature 1: numeric vector
  BRC1_UP_DE_geneList_FC = BRC1_UP_DE_list$logFC_GFP_BRC1ind_vs_brc1
  
  ## feature 2: named vector
  names(BRC1_UP_DE_geneList_FC) = as.character(row.names(BRC1_UP_DE_list))
  
  ## feature 3: sorted vector
  BRC1_UP_DE_geneList_FC = sort(BRC1_UP_DE_geneList_FC, decreasing = TRUE)
```

### GO over-representation analysis of UP genes in BRC1ind
```{r}
# Test which biological processes are over-represented in BF DT of BRC1

UP_DE_BRC1_enrichedGO_250max =  enrichGO(gene          = row.names(BRC1_UP_DE_list),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = 250,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

UP_DE_BRC1_enrichedGO_500max =  enrichGO(gene          = row.names(BRC1_UP_DE_list),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = 500,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

UP_DE_BRC1_enrichedGO_1000max = enrichGO(gene          = row.names(BRC1_UP_DE_list),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = 1000,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

UP_DE_BRC1_enrichedGO_Infmax = enrichGO(gene          = row.names(BRC1_UP_DE_list),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = Inf,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

# Save filtered GO terms in a table
UP_DE_BRC1_enrichedGO_Infmax@result %>% filter(p.adjust<0.05) %>% write.table(.,
            file = "./R_output/GO_terms_enrichment_analysis/BRC1_UP_DE_list_GO_terms_InfmaxGSS.txt",
            row.names = F, sep = "\t", quote = F, na = "")
```

### Visualization of enriched GO terms
```{r}
#goplot(BF_DT_BRC1_enrichedGO) # directed acyclic graph
UP_DE_BRC1_cnetplot_250maxGSS = cnetplot(UP_DE_BRC1_enrichedGO_250max, categorySize="p.adjust",
                                   foldChange=BRC1_UP_DE_geneList_FC) # category netplot with FC values
UP_DE_BRC1_cnetplot_500maxGSS = cnetplot(UP_DE_BRC1_enrichedGO_500max, categorySize="p.adjust",
                                   foldChange=BRC1_UP_DE_geneList_FC) # category netplot with FC values
UP_DE_BRC1_cnetplot_1000maxGSS = cnetplot(UP_DE_BRC1_enrichedGO_1000max, categorySize="p.adjust", 
                                    foldChange=BRC1_UP_DE_geneList_FC) # category netplot with FC values

# Saving Gene-Concept Networks
png("./R_output/GO_terms_enrichment_analysis/BRC1ind_UP_DE_Gene-Concept_Net_250maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
UP_DE_BRC1_cnetplot_250maxGSS
dev.off()

png("./R_output/GO_terms_enrichment_analysis/BRC1ind_UP_DE_Gene-Concept_Net_500maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
UP_DE_BRC1_cnetplot_500maxGSS
dev.off()

png("./R_output/GO_terms_enrichment_analysis/BRC1ind_UP_DE_Gene-Concept_Net_1000maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
UP_DE_BRC1_cnetplot_1000maxGSS
dev.off()
```


# GFP:BRC1ind DN DE genes #
```{r}
BRC1_DN_DE_list = BRC1_DE_list %>% filter(logFC_GFP_BRC1ind_vs_brc1<1)

#Creating a geneList with -log10(FC) values of genes with FDR < 0.05
  ## feature 1: numeric vector
  BRC1_DN_DE_geneList_FC = BRC1_DN_DE_list$logFC_GFP_BRC1ind_vs_brc1
  
  ## feature 2: named vector
  names(BRC1_DN_DE_geneList_FC) = as.character(row.names(BRC1_DN_DE_list))
  
  ## feature 3: sorted vector
  BRC1_DN_DE_geneList_FC = sort(BRC1_DN_DE_geneList_FC, decreasing = TRUE)
```

### GO over-representation analysis of DN genes in BRC1ind
```{r}
# Test which biological processes are over-represented in BF DT of BRC1

DN_DE_BRC1_enrichedGO_250max =  enrichGO(gene          = row.names(BRC1_DN_DE_list),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = 250,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

DN_DE_BRC1_enrichedGO_500max =  enrichGO(gene          = row.names(BRC1_DN_DE_list),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = 500,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

DN_DE_BRC1_enrichedGO_1000max = enrichGO(gene          = row.names(BRC1_DN_DE_list),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = 1000,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

DN_DE_BRC1_enrichedGO_Infmax = enrichGO(gene          = row.names(BRC1_DN_DE_list),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = Inf,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

# Save filtered GO terms in a table
DN_DE_BRC1_enrichedGO_Infmax@result %>% filter(p.adjust<0.05) %>% write.table(.,
            file = "./R_output/GO_terms_enrichment_analysis/BRC1_DN_DE_list_GO_terms_InfmaxGSS.txt",
            row.names = F, sep = "\t", quote = F, na = "")
```

### Visualization of enriched GO terms
```{r}
#goplot(BF_DT_BRC1_enrichedGO) # directed acyclic graph
DN_DE_BRC1_cnetplot_250maxGSS = cnetplot(DN_DE_BRC1_enrichedGO_250max, categorySize="p.adjust",
                                   foldChange=BRC1_DN_DE_geneList_FC) # category netplot with FC values
DN_DE_BRC1_cnetplot_500maxGSS = cnetplot(DN_DE_BRC1_enrichedGO_500max, categorySize="p.adjust",
                                   foldChange=BRC1_DN_DE_geneList_FC) # category netplot with FC values
DN_DE_BRC1_cnetplot_1000maxGSS = cnetplot(DN_DE_BRC1_enrichedGO_1000max, categorySize="p.adjust", 
                                    foldChange=BRC1_DN_DE_geneList_FC) # category netplot with FC values

# Saving Gene-Concept Networks
png("./R_output/GO_terms_enrichment_analysis/BRC1ind_DN_DE_Gene-Concept_Net_250maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
DN_DE_BRC1_cnetplot_250maxGSS
dev.off()

png("./R_output/GO_terms_enrichment_analysis/BRC1ind_DN_DE_Gene-Concept_Net_500maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
DN_DE_BRC1_cnetplot_500maxGSS
dev.off()

png("./R_output/GO_terms_enrichment_analysis/BRC1ind_DN_DE_Gene-Concept_Net_1000maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
DN_DE_BRC1_cnetplot_1000maxGSS
dev.off()
```



# BRC1 BF DT #
```{r}
row.names(BRC1_BF_DT) = BRC1_BF_DT$Geneid

#Creating a geneList with -log10(FC) values of genes with FDR < 0.05
  ## feature 1: numeric vector
  BRC1_BF_DT_geneList_FC = BRC1_BF_DT$logFC_GFP_BRC1ind_vs_brc1
  
  ## feature 2: named vector
  names(BRC1_BF_DT_geneList_FC) = as.character(row.names(BRC1_BF_DT))
  
  ## feature 3: sorted vector
  BRC1_BF_DT_geneList_FC = sort(BRC1_BF_DT_geneList_FC, decreasing = TRUE)
```

### GO over-representation analysis of BF DT of BRC1
```{r}
# Test which biological processes are over-represented in BF DT of BRC1

BF_DT_BRC1_enrichedGO_500max = enrichGO(gene          = row.names(BRC1_BF_DT),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = 500,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

BF_DT_BRC1_enrichedGO_1000max = enrichGO(gene          = row.names(BRC1_BF_DT),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = 1000,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

BF_DT_BRC1_enrichedGO_Infmax = enrichGO(gene          = row.names(BRC1_BF_DT),
                                         OrgDb         = org.At.tair.db,
                                         keyType       = "TAIR",
                                         ont           = "BP",
                                         minGSSize     = 10,
                                         maxGSSize     = Inf,
                                         pAdjustMethod = "BH",
                                         pvalueCutoff  = 0.05,
                                         qvalueCutoff  = 1,
                                         readable      = TRUE) #%>% simplify()

# Save filtered GO terms in a table
BF_DT_BRC1_enrichedGO_Infmax@result %>% filter(p.adjust<0.05) %>% write.table(.,
            file = "./R_output/GO_terms_enrichment_analysis/BRC1_BF_DT_GO_terms_InfmaxGSS.txt",
            row.names = F, sep = "\t", quote = F, na = "")
```

### Visualization of enriched GO terms
```{r}
#goplot(BF_DT_BRC1_enrichedGO) # directed acyclic graph
BRC1_cnetplot_500maxGSS = cnetplot(BF_DT_BRC1_enrichedGO_500max, categorySize="p.adjust",
                                   foldChange=BRC1_BF_DT_geneList_FC) # category netplot with FC values
BRC1_cnetplot_1000maxGSS = cnetplot(BF_DT_BRC1_enrichedGO_1000max, categorySize="p.adjust", 
                                    foldChange=BRC1_BF_DT_geneList_FC) # category netplot with FC values

# Saving Gene-Concept Networks
png("./R_output/GO_terms_enrichment_analysis/BRC1ind_BF_DT_Gene-Concept_Net_500maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
BRC1_cnetplot_500maxGSS
dev.off()

png("./R_output/GO_terms_enrichment_analysis/BRC1ind_BF_DT_Gene-Concept_Net_1000maxGSS.png", width = 30, height = 20, units = "cm", res = 300)
BRC1_cnetplot_1000maxGSS
dev.off()
```
